Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-08-10 docs/manuscript/paper.tex
example likelihood surfaces

\subsubsection{Measurement equation}
\textcolor{red}{TK: Does this subsection needs a better name?} \newline 
\noindent \textcolor{red}{TK: this subsection has been overhauled per comments from AM. The structure is now: \newline 1. Radio pulses described by general geometric object $p$ \newline 2. Shift in $p$ due to a GW (no derivation, just definition of terms) \newline 3.Relate shift in $p$ to a shift in $f$ by clarifying that $u=(1,0,0,0)$ for PSR and observer. \newline 4. Give final measurement equation} \newline

In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave. They are described by the geometric object $\vec{p}$ which we can identify  as the 4-momentum of the radio photon. The presence of a GW induces a shift in the temporal component of the covariant 4-momentum between the emitter and the observer, i.e. $\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $, as \citep[e.g.][]{Maggiore}
\begin{equation}
 \Delta p_t = \frac{\omega}{2} \frac{ h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}) }  \left(1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right)
	\label{eq:momentum_shift}
\end{equation}
where $\omega$ is the angular pulse frequency $(= 2 \pi f_{\rm p})$ measured in the momentarily comoving reference frame (MCRF) of an observer, $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar. We take the pulsar location to be constant i.e.  $\boldsymbol{q}$ is not a function of time. In practice the pulsar locations vary with respect to the Earth but are constant with respect to the SSB. This barycentering correction is typically applied when generating TOAs using e.g. {\sc tempo2} and related timing software. Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\vec{u}$ is given by the coordinate-independent expression:
\begin{equation}
	f = p_{\alpha} u^{\alpha} \ . 
	\label{eq:freq_temporal}
\end{equation}
Due to the kinematical corrections from the barycentering process, 
\begin{equation}
	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
\end{equation}
where the perturbations to $\vec{u}$ from the GW are of higher order and neglected. In this case the shift in the momentum from Equation \ref{eq:momentum_shift} can be related to a shift in the frequency as,
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}(t)
	\label{eq:measurement}
\end{equation}
where \textcolor{red}{TK: what is the best notation here to give $q^i$ an $(n)$ superscript?}
\begin{equation}
	g^{(n)}(t) = 1 -  \frac{1}{2} \frac{h_{ij} (t; \boldsymbol{x}= 0)q^{i}_{(n)} q^{j}_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \left(1 -e^{i \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}}\right) \ .
	\label{eq:g_func}
\end{equation}
It is will also prove instructive to express Eq. \ref{eq:g_func} in a trigonometric form as,
or in a trigonometric form 
\begin{align}
	g^{(n)}(t) &= 1 - \frac{1}{2} \frac{ H_{ij}q^i_{(n)} q^j_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) } \nonumber \\
	& \times \left[ \cos(-\Omega t +\Phi_0) - \cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \right) \right] \ .
	\label{eq:g_func_trig}
\end{align}
Equations \ref{eq:measurement}, \ref{eq:g_func_trig}, \ref{eq:g_func}  define a measurement equation that relates the intrinsic pulsar spin frequency to the radio pulse frequency measured by an observer on Earth.

@@

@@

\subsubsection{Multiple noise realisations} \label{sec:multiple_noise}

The above discussion was just for a single realisation of the noise processes. It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim}, \ref{sec:detection} is not particularly advantageous or convenient for our method.

To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate 1000 realisations of the process noise and measurement noise.

For each realisation of the data we can then independently estimate the static parameters $\boldsymbol{\theta}$.

The estimated values of $\boldsymbol{\theta}_{\rm gw}$ for 5 realisations of the noise are shown in Fig \ref{fig:corner_plot_2} which is analogous to Fig \ref{fig:corner_plot_1} but now shows multiple realisations of the noise.

We can see that the results agree well with the case of a single noise realisation. The estimates of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are highly consistent across the different realisations, agreeing with the injected values with high fidelity.

The variance in the estimates of $\iota$ and $h_0$ is much greater than for the other parameters.

This is due to the nested sampler effectively trading off accuracy in one parameter against the other; from Eq \ref{eq:hphx} \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$.  For example, from Eq \ref{eq:hphx2} it is not clear if a larger $h_{\times}$ is due to a larger strain $h$ or a different inclination $\iota$. This degeneracy is broken by the relation for $h_{+}$ which does separate $h$ and $\iota$

Whilst both parameters are then generally identifiable, the specific algebraic relation between the two and how these parameters get mapped into the measurement equation makes it difficult for our algorithm to estimate both simultaneously, instead typically trading off one parameter against the other.

If we fix one of these parameters, it is then possible for our approach to recover the other parameter with much higher accuracy. This is demonstrated in Fig X which is analogous to the representative exxample, but now we have fixed $h$ at its true value (i.e. reduced the prior to a delta function). WE can see that the method can now infer $\iota$ which much greater accuracy.

Similar to the estimates of $\boldsymbol{\theta}_{\rm gw}$, the estimates of $\boldsymbol{\theta}_{\rm psr}$ (c.f.  Fig \ref{fig:example_psr_params}) show a consistency between the single and multiple noise realisations though for brevity we do not show the results here \textcolor{red}{TK: is this fair? Don't want to clog up paper with extra figures that don't offer any real insight.}

In the case where h is fixed...

whilst the estimates on  are much more varied.

In order to visualise the 1000 realisations, it is convenient to consider the distribution in the median of the posterior that is inferred for each static parameter, for each realisation of the noise. The resulting distributions are presented in Fig \ref{fig:median_distriubutins}

The bias that was evident for some of the parameters

  \textcolor{red}{with the estimate on iota}

are entirely consistent with the single

, which is analogous to Fig Y but with multiple  now shown.

\begin{figure*}
	\includegraphics[width=0.8\textwidth, height =0.8\textwidth ]{images/stacked_GW_plot_1000}
	\caption{}
	\label{fig:corner_plot_2}
\end{figure*}

\begin{figure*}
	\setkeys{Gin}{width=\linewidth}   
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics{images/distribution_omega_gw}
		\caption{$\Omega$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics{images/distribution_phi0_gw}
		\caption{$\Phi_0$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics{images/distribution_psi_gw}
		\caption{$\psi$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics{images/distribution_iota_gw}
		\caption{$\iota$}
	\end{subfigure}

	\medskip
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics{images/distribution_delta_gw}
		\caption{$\delta$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics{images/distribution_alpha_gw}
		\caption{$\alpha$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.22\textwidth}
		\includegraphics{images/distribution_h}
		\caption{$h$}
	\end{subfigure}

	\caption{some caption \textcolor{red}{TK: these plots are ugly. Is there a better way to show this?}}\label{fig:median_distriubutins}
\end{figure*}

@@

The above was for a particular SBHB system

can we also get a sky map using the KF? 
\subsection{Bias by dropping pulsar terms}

@@

single pulsar performance.

hellings downs

multiple sources

spherical localisatin plot would be cool e.g.

\url{https://github.com/stephengreen/lfi-gw}

comparison with alternative methods

more stringent priors on f and fdot 
Points to discuss:

\begin{itemize}
	\item Non constant time sampling? Different pulsars sampled at different times...
	\item pulsar terms obv
\end{itemize}

@@

Our method is not contingent on the choice of PTA, and naturally extend to other PTAs or PTAs with more pulsars.

Questions

\begin{itemize}
	\item PTAs like MSPs for small timing noise. Can we get away with large timing noise, and use more pulsars? Some pulsars are more useful than others, e.g. arXiv 2211.03201
	\item Can we also estimate radiometer noise? Is this useful?
\end{itemize}

@@

>>>>

\subsubsection{Measurement equation}
In general radio pulses from a pulsar are transmitted as amplitude modulations of a radio-frequency carrier wave. They are described by the geometric object $\vec{p}$ which we can identify  as the 4-momentum of the radio photon. The presence of a GW induces a shift in the temporal component of the covariant 4-momentum between the emitter and the observer, i.e. $\Delta p_t = p_t|_{\rm observer} - p_t|_{\rm emitter} $, as \citep[e.g.][]{Maggiore}
\begin{equation}
 \Delta p_t = \frac{\omega}{2} \frac{ h_{ij} (t; \boldsymbol{x}= 0)q^i q^j }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}) }  \left(1 -e^{i \Omega (1 + \boldsymbol{n}\cdot \boldsymbol{q})  d}\right)
	\label{eq:momentum_shift}
\end{equation}
where $\omega$ is the angular pulse frequency $(= 2 \pi f_{\rm p})$ measured in the momentarily comoving reference frame (MCRF) of an observer, $h_{ij} = g_{ij} - \eta_{ij}$, $\boldsymbol{q}$ is the unit vector connecting the observer and the pulsar and $d$ is the distance to the pulsar. We take the pulsar location to be constant i.e.  $\boldsymbol{q}$ is not a function of time. In practice the pulsar locations vary with respect to the Earth but are constant with respect to the SSB. This barycentering correction is typically applied when generating TOAs using e.g. {\sc tempo2} and related timing software. Generally the measured frequency of a photon recorded by an observer who is travelling with 4-velocity $\vec{u}$ is given by the coordinate-independent expression:
\begin{equation}
	f = p_{\alpha} u^{\alpha} \ . 
	\label{eq:freq_temporal}
\end{equation}
Due to the kinematical corrections from the barycentering process, 
\begin{equation}
	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
\end{equation}
where the perturbations to $\vec{u}$ from the GW are of higher order and neglected. In this case the shift in the momentum from Equation \ref{eq:momentum_shift} can be related to a shift in the frequency as,
\begin{equation}
	f_{\rm m}^{(n)}(t) = f_{\rm p}^{(n)}(t-d) g^{(n)}(t)
	\label{eq:measurement}
\end{equation}
where \textcolor{red}{TK: what is the best notation here to give $q^i$ an $(n)$ superscript?}
\begin{equation}
	g^{(n)}(t) = 1 -  \frac{1}{2} \frac{h_{ij} (t; \boldsymbol{x}= 0)q^{i}_{(n)} q^{j}_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) }  \left(1 -e^{i \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)}}\right) \ .
	\label{eq:g_func}
\end{equation}
It is will also prove instructive to express Eq. \ref{eq:g_func} in a trigonometric form as,
or in a trigonometric form 
\begin{align}
	g^{(n)}(t) &= 1 - \frac{1}{2} \frac{ H_{ij}q^i_{(n)} q^j_{(n)} }{(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)}) } \nonumber \\
	& \times \left[ \cos(-\Omega t +\Phi_0) - \cos \left(-\Omega t +\Phi_0 + \Omega \left(1 + \boldsymbol{n}\cdot \boldsymbol{q}^{(n)} \right)  d^{(n)} \right) \right] \ .
	\label{eq:g_func_trig}
\end{align}
Equations \ref{eq:measurement}, \ref{eq:g_func_trig}, \ref{eq:g_func}  define a measurement equation that relates the intrinsic pulsar spin frequency to the radio pulse frequency measured by an observer on Earth.

@@

In practice looser, leave them loose to subject method to more stringent test i.e. can we still estimate GW parameters without highly accurate t0 frequencies.

@@

\subsubsection{Multiple noise realisations} \label{sec:multiple_noise}
\begin{figure*}
	\includegraphics[width=0.8\textwidth, height =0.8\textwidth ]{images/stacked_GW_plot_1000}
	\caption{}
	\label{fig:corner_plot_2}
\end{figure*}
The above discussion was just for a single realisation of the noise processes. It is important to confirm that our method is robust against different realisations of the noise and that the specific realisation of the noisy data used in Sections \ref{sec:parameter_estim}, \ref{sec:detection} is not particularly advantageous or convenient for our method. To this end we take our representative example from Table \ref{tab:parameters_and_priors} and generate 1000 realisations of the process noise and measurement noise. For each realisation of the data we can then independently estimate the static parameters, $\boldsymbol{\theta}$. In Figure \ref{fig:corner_plot_2} we show the results for the  estimated values of $\boldsymbol{\theta}_{\rm gw}$ for 9 realisations of the noise. This figure is analogous to the single noise realisation of Figure \ref{fig:corner_plot_1}. We plot only 9 realisations of the noise rather than the full set of 1000 realisations so as to not overcrowd the figure. We can see from Figure \ref{fig:corner_plot_2} that the results for multiple noise realisations agree well with the results that we saw in  Figure \ref{fig:corner_plot_1} for a single noise realisation. The estimates of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are highly consistent across the different noise realisations i.e. the variance between the different marginalized posteriors is generally low. The estimates generally agree with the injected values with high fidelity. Conversely,for the parameters $\iota$ and $h_0$ the variance between the different marginalized posteriors is generally high; we infer highly different estimates of these parameters depending on the particular realisation of the noise. \newline

 The large degree of variance in the estimates of $\iota$ and $h_0$ is due to the nested sampling algorithm effectively trading off accuracy in one parameter against the accuracy in other. From Eqs. \ref{eq:hphx} \ref{eq:hphx2} we can see that there is a weak degeneracy between $\iota$ and $h$. For example, from Eq \ref{eq:hphx2} it is not clear if a larger value of the cross polarisation strain $h_{\times}$ is due to the system having a larger strain amplitude $h_0$, or a different inclination $\iota$. This degeneracy is broken generally by the relation for $h_{+}$, Eq \ref{eq:hphx}. Whilst this ensures that both parameters are then generally identifiable formally, in practice an identifiability issue persists since the log-likelihood surface in $\iota -h_0$ space plateaus across a broad range of reasonable values (\textcolor{red}{TK: Figure here would be useful}).  Consequently it is difficult for our algorithm to select a particular point in this plateau. If we fix one of the parameters at its true value, taking a slice through the likelihood surface, it is then possible to identify a particular point in the plateau surface and the variance in the estimates between different noise realisations decreases. This is demonstrated in Figure \ref{fig:iota}

As we will see in Section \ref{sec:bias}

 larger strain or

 from

 If we fix one of these parameters, it is then possible for our approach to recover the other parameter with much higher accuracy. This is demonstrated in 
for the inferred posteriors of $\iota$ for the same 5 noise realisations of Fig \ref{fig:corner_plot_2}, but now taking $h$ as a known parameter fixed at its true value.

\begin{figure*}

		\begin{subfigure}[b]{0.48\textwidth}
		\includegraphics[width=\textwidth]{images/stacked_GW_plot_iota_h}
		\caption{}
		\label{fig:example1_iota}
	\end{subfigure} 
	\hfill
	\begin{subfigure}[b]{0.48\textwidth}
		\includegraphics[width=\textwidth]{images/stacked_GW_plot_iota_h_fixed}
		\caption{}
		\label{fig:example2_iota}
	\end{subfigure}

	\caption{Figure showing distribution of $\iota$ for a known/fixed $h$}
	\label{fig:iota}
\end{figure*}
We can see... \newline

Note that whilst the variance between the different noise realisations has decreased, the discrepancy between the estimated value and the true recovered value has increased.

Similar to the estimates of $\boldsymbol{\theta}_{\rm gw}$, the estimates of $\boldsymbol{\theta}_{\rm psr}$ (c.f.  Fig \ref{fig:example_psr_params}) show a consistency across multiple noise realisations, though for brevity we do not show the results here \textcolor{red}{TK: is this fair? Don't want to clog up paper with extra figures that don't offer any real insight.} \newline

For each realisation of the noise we infer a marginalised 1-D posterior for each parameter. Each posterior has a median value. We plot the distribution of these medians over 1000 noise realisations for each parameter of $\boldsymbol{\theta}_{\rm gw}$ in Figure \ref{fig:median_distriubutins}. For each of these distributions over the noise realisations we can also calculate the median value (i.e. the median of the medians) and compare it with the true injection value. This is shown by the red and orange dashed lines respectively in the Figure.
\begin{figure*}
	\setkeys{Gin}{width=\linewidth}

	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/distribution_omega_gw}
		\caption{$\Omega$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/distribution_phi0_gw}
		\caption{$\Phi_0$}
	\end{subfigure}
	\hfill	
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/distribution_psi_gw}
		\caption{$\psi$}
	\end{subfigure}
	\medskip

		\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/distribution_iota_gw}
		\caption{$\iota$}
	\end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/distribution_delta_gw}
		\caption{$\delta$}
	\end{subfigure}
	\hfill	
	\begin{subfigure}[b]{0.3\textwidth}
		\includegraphics[width=\textwidth]{images/distribution_alpha_gw}
		\caption{$\alpha$}
	\end{subfigure}
	\medskip

	\medskip
	\begin{subfigure}[b]{0.33\textwidth}
		\includegraphics{images/distribution_h}
		\caption{$h$}
	\end{subfigure}

	\caption{some caption \textcolor{red}{TK: these plots are ugly. Is there a better way to show this?}}\label{fig:median_distriubutins}
\end{figure*}
Analogous to the results we saw in Figure \ref{fig:corner_plot_2}, the distributions of the median values of the parameters $\Omega, \Phi_0, \psi, \delta$ and $\alpha$ are very narrow with a generally small discrepancy between the inferred value and the injected value. The parameters $\psi$ and $\alpha$ do exhibit a small bias of magnitude $\sim 0.1$ and $\sim 0.04$ respectively which we will discuss in Section \ref{sec:bias}.

@@

\textcolor{red}{TK: TBD}

\subsection{Bias by dropping pulsar terms}\label{sec:bias}

@@

Some potential further points for discussion

\begin{itemize}
	\item What is the performance of a single pulsar?
	\item How does this map to Hellings-downs?
	\item How does this map to having more than one source? Or a chrmomatic source?
	\item Non constant time sampling?
	\item Any further discussion on pulsar terms?
	\item PTAs like MSPs for small timing noise. Can we get away with large timing noise, and use more pulsars? Some pulsars are more useful than others, e.g. arXiv 2211.03201
	\item Can we also estimate radiometer noise? Is this useful?
\end{itemize}

@@

\textcolor{red}{TK: TBD}

@@

\textcolor{red}{TK: This section needs cleaning up}
